---
Title: Data Mining Lab 1
Authors: Max Link, Logan Lu, Jadon Klipsch
Date: 2025-02-21
Description: In this project, we will focus on cleaning and understanding the data. We will follow the Crisp-DM framework 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r} 
# Load required libraries
library(dplyr)      # For data manipulation
library(ggplot2)    # For visualizations
library(tidyr)      # For cleaning data
```
```{r dataframe}
# reading in Mobility Report
data <- read.csv('COVID-19/Global_Mobility_Report.csv')

# displaying first 20 rows of report
head(data, 20)
```

Crisp-DM Framework:
1. Problem Description (Business Understanding) [10]
Describe the Problem: What is COVID-19, and what is social distancing and flattening the
curve? Why is it important to look at data about the virus spread, hospitalizations, and
available resources? [3 points]

COVID-19 is a highly contagious respiratory illness caused by the SARS-CoV-2 virus, first identified in late 2019. Social distancing involves reducing physical interactions to slow the virus’s spread, aiming to "flatten the curve"—i.e., reduce the peak number of cases to avoid overwhelming healthcare systems. Analyzing data on virus spread, hospitalizations, and resources is critical to understand transmission patterns, predict healthcare demands, and develop effective interventions (e.g., vaccines or supplies by Johnson & Johnson). 

Choose a stakeholder for whom you analyze and, later on, model the data. Define some
questions that are important for this stakeholder. What decisions can your stakeholders make, and how would they affect COVID-19 outcomes? Brainstorm this a lot since this choice will guide your exploration of this and all the following projects. Make sure you can produce actionable recommendations for these questions using your data later in your report. [7 point]

Johnson & Johnson (J&J), a medical company focused on pharmaceuticals, medical devices, and consumer health products.

Questions for J&J:

How does population density affect COVID-19 transmission rates across U.S. counties?

What is the relationship between social distancing compliance (mobility changes) and case growth?

How do hospital resources (e.g., beds per capita) correlate with mortality rates?

Which counties showed delayed social distancing responses, and how did this impact case surges?

What demographic or geographic factors predict higher demand for medical supplies or vaccines? 

Johson and Johnson can make the following decisions: 
- Prioritize vaccine or medical supply distribution to high-risk, densely populated areas.

- Partner with counties lacking resources to provide equipment or therapeutics.

- Adjust production timelines based on predicted surges linked to mobility patterns.


 2. Data Understanding [45]
 You must include all three provided datasets in your analysis!

Describe what data is available. Choose 5-10 important variables for the questions you have identified in the section above. Describe the type of data (scale, values, etc.) of the most critical variables in the data. [9 point]


Verify data quality: Are there missing values? Duplicate Data? Outliers? Are those mistakes? How can these be fixed? Ensure your report states how much data is removed and how much you have left. [9 Points]


Give appropriate statistics (range, mode, mean, median, variance, etc.) for the most important variables in these files and describe what they mean or if you find something interesting. [9 points]


Visually explore the chosen attributes appropriately. Provide an interpretation for each graph. Explain why you chose the visualization for each attribute type. [9 points]


Explore relationships between attributes: Look at the attributes and then use cross-tabulation, correlation, group-wise averages, box plots, etc., as appropriate. [9 points]


3. Data Preparation [30]
Create a data set with objects as rows and features as columns. Use as objects counties in the US. The data set must be included in your report. Provide a table in your Word document that shows the data (values) for the first 10 rows for all the features you have selected and/or created. Interesting additional features may be, for example: [30 points]


When was the first case reported?
How (densely) populated is the county?
What resources does a county have (money, hospital)?
What is the social distancing response, and how long did it take after the first case?
You can come up with more critical questions for your chosen stakeholder.


4. Data Preparation [15]
Formulate some recommendations for the questions developed in section 1. based on the results in 2. and 3. Make sure your recommendations are based on data and are actionable for the stakeholder (i.e., the stakeholder has the power to execute the recommendations). [15 points]
